---
output: 
  github_document
bibliography: inst\\REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  eval = TRUE
)
```

# irpeat

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

irpeat is an R package that contains simple functions to analyse infrared spectra of peat samples. Some functions may also work with organic matter samples in general.

Provided functions for analyzing infrared spectra of peat are:

1. Computation of several humification indices.
2. Klason lignin mass fraction (following @Hodgkins.2018).
3. Holocellulose mass fraction (following @Hodgkins.2018).
4. Peat electron accepting capacity (following @Teickner.2022).
5. Peat electron donating capacity (following @Teickner.2022).

### How to install

You can install irpeat from GitHub using R via:
```{r installation, eval = FALSE}
remotes::install_github(repo = "henningte/irpeat")
```

irpeat relies on the R package [ir](https://github.com/henningte/ir) for handling infrared spectra.

### How to use

You can load irpeat in R with:

```{r load_irpeat}
library(irpeat)

# load additional packages needed for this tutorial
library(ir)
library(ggplot2)
library(magrittr)
```

You can test irpeat with sample data from the R package ir:

```{r ir_sample_data_load}
ir::ir_sample_data
```

`ir::ir_sample_data` contains various ATR-MIR spectra of organic reference material (e.g. newspaper, wood, grass).

A simple workflow could be, for example, to baseline correct the spectra (using functions of the package ir) compute various humification indices and Klason lignin and holocellulose mass fractions in the samples. We use only the first few spectra from `ir::ir_sample_data` to speed the computations a bit up.

```{r sample_data_workflow}
x <- 
  ir::ir_sample_data[1:10, ] %>%                                # data
  ir::ir_bc(method = "rubberband") %>%                          # baseline correction
  irpeat::irp_hi() %>%                                          # humification indices
  irpeat::irp_content(variable = c("klason_lignin_hodgkins",    # Klason lignin and
                                   "holocellulose_hodgkins"))   # holocellulose content
```

`x` is identical to `ir::ir_sample_data[1:10, ]`, but contains additional columns for the computed humification indices (h1, h2, h3, h4) and the computed Klason lignin content (klason_lignin_hodgkins) and holocellulose content (holocellulose_hodgkins).

```{r x_inspect}
x
```

Plot of the humification index (ratio of the intensities at 1420 and 1090 cm<sup>-1</sup> [@Broder.2012]) versus the Klason lignin content:

```{r x_plot}
ggplot2::ggplot(x, aes(x = quantities::drop_quantities(klason_lignin_hodgkins) * 100, y = hi1)) + 
  ggplot2::geom_point() +
  ggplot2::labs(x = "Klason lignin content [mass-%]", 
                y = expression("Ratio of the intensities at"~1420~and~1090~cm^{-1}))

```


All computed quantities come with units and standard errors (thanks to the [quantities](https://github.com/r-quantities/quantities) package):

```{r x_klh_inspect_quantities}
x$holocellulose_hodgkins
x$klason_lignin_hodgkins
```

### Future development

Henning Teickner plans, as part of his PhD project, to extensively extent irpeat by developing a set of calibration models that can predict various peat physicochemical properties from mid infrared spectra. These models should be finished by November 2022. Currently, a data compendium ([pmird](https://henningte.github.io/pmird/index.html)) is developed to collect the data required for this task.

### How to cite

Please cite this R package as:

> Henning Teickner, Suzanne B. Hodgkins (`r format(Sys.Date(), "%Y")`). _irpeat: Functions to Analyze Mid Infrared Spectra of Peat Samples_. Accessed `r format(Sys.Date(), "%Y-%m-%d")`. Online at <https://github.com/henningte/irpeat>.


### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) attribution requested in reuse. See the sources section for data sources and how to give credit to the original author(s) and the source.

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### Sources

The data and prediction models for holocellulose and Klason are derived from @Hodgkins.2018 and were restructured to match the requirements of ir. The original article containing the data can be downloaded from https://www.nature.com/articles/s41467-018-06050-2 and is distributed under the Creative Commons Attribution 4.0 International License (http://creativecommons.org/licenses/by/4.0/). The data on Klason lignin and holocellulose content was originally derived from @LaCruz.2016.

Data and models for the electrochemical accepting and donating capacities (EAC, EDC) of peat were derived from @Teickner.2022 and @Teickner.2021c

This packages was developed in R (`r R.Version()$version.string`) [@RCoreTeam.2019] using functions from devtools [@Wickham.2019], usethis [@Wickham.2019b], rrtools [@Marwick.2019] and roxygen2 [@Wickham.2019c].

### References
